ANCOVA Comparison Simulations: Gaussian Data

### 2017-02-15 ms/SBR-template.tex
Progress on suppl files

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf  A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
% 200 words or less
Hypothesis tests based on linear models are widely accepted by organizations that regulate clinical trials.
These tests are derived using strong assumptions about the data-generating process so that the resulting inference is based on parametric distributions.
Because these methods are well understood and robust, they are sometimes applied to data that depart from assumptions, such as ordinal integer scores.
Permutation tests are a nonparametric alternative that require minimal assumptions which are often guaranteed by the randomization that was conducted.
We compare analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification, to several permutation tests based on linear models that control for pretreatment covariates.
In simulations using a variety of data-generating processes, some of which violate the parametric assumptions,
the permutation tests maintain power comparable to ANCOVA.
We illustrate the use of these permutation tests alongside ANCOVA with data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
Given the considerable costs and scientific importance of clinical trials, one may want to include an additional robust method, such as a linear model permutation test, as a check on the statistical inference for the main study endpoints.
\end{abstract}

\noindent%
{\it Keywords:}  nonparametric, linear model, ANCOVA, inference
\vfill

@@

Our second set of simulations used the same design as the previous ones, but modified the sizes of the treatment groups.
Within each stratum, 4 individuals received one treatment and 12 received the other.
This mimics some real-world experiments, where the more expensive treatment is administered less frequently than the placebo or standard of care.
It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
Here, the data had equal variances so we would expect power, not the test level, to be an issue.
(See the supplementary Gaussian simulations for simulations using data with heterogeneous variances.)
Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
In this case, the permutation tests (except for the simple stratified permutation) had slightly higher power than ANCOVA, but the difference was not substantial.
%\begin{figure}
%\centering
%\includegraphics[width = \textwidth]{fig/imbalanced_simulation_power}
%\caption{Empirical power curves for the Gaussian simulated data with imbalanced group sizes}
%\label{fig:imbalanced_sim_power}
%\end{figure}
%\begin{center}
%\input{fig/imbalanced_simulation_power_summary.tex}
%\end{center}

@@

\bigskip
\begin{center}
\todo{}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

@@

\item[Title:] Brief description. (file type)

>>>>

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf  A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Experiments}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
% 200 words or less
Hypothesis tests based on linear models are widely accepted by organizations that regulate clinical trials.
These tests are derived using strong assumptions about the data-generating process so that the resulting inference can be based on parametric distributions.
Because these methods are well understood and robust, they are sometimes applied to data that depart from assumptions, such as ordinal integer scores.
Permutation tests are a nonparametric alternative that require minimal assumptions which are often guaranteed by the randomization that was conducted.
We compare analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification, to several permutation tests based on linear models that control for pretreatment covariates.
In simulations using a variety of data-generating processes, some of which violate the parametric assumptions,
the permutation tests maintain power comparable to ANCOVA.
We illustrate the use of these permutation tests alongside ANCOVA with data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
Given the considerable costs and scientific importance of clinical trials, one may want to include an additional nonparametric method, such as a linear model permutation test, as a robustness check on the statistical inference for the main study endpoints.
\end{abstract}

\noindent%
{\it Keywords:}  Nonparametric methods, linear model, analysis of covariance, analysis of designed experiments
\vfill

@@

Our second set of simulations used the same design as the previous ones, but modified the sizes of the treatment groups.
Within each stratum, 4 individuals received one treatment and 12 received the other.
This mimics some real-world experiments, where the more expensive treatment is administered less frequently than the placebo or standard of care.
It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
Here, the data had equal variances so we would expect power, not the test level, to be an issue.
(See the supplementary Gaussian simulations for simulations using data with heterogeneous variances.)
Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
In this case, the permutation tests (except for the simple stratified permutation) had slightly higher power than ANCOVA, but the difference was not substantial.

%\begin{center}
%\input{fig/imbalanced_simulation_power_summary.tex}
%\end{center}

@@

\bigskip
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}
All files are also available at \url{https://github.com/kellieotto/ancova-permutations}.

@@

\item[Gaussian data simulations:] Simulations using Gaussian data in Section~\ref{sec:simulations}, including code and results. (PDF)
\item[Imbalanced design simulations:] Simulations using data with different stratum sizes in Section~\ref{sec:simulations}, including code and results. (PDF)
\item[Skewed data simulations:] Simulations using discrete, skewed data in Section~\ref{sec:simulations}, including code and results. (PDF)
\item[Clinical trial data:] Dataset used in Section~\ref{sec:results} to compare methods. (csv)
\item[README:] Data descriptor file. (txt)
\item[Results:] Detailed explanation, code, and results of comparing methods using the clinical trial dataset in Section~\ref{sec:results}. (PDF)

### 2017-02-15 simulations/simulation_gaussian.Rmd
Progress on suppl files

plot_power_curves <- function(power_mat, title){
  melt(power_mat) %>% 
  mutate("pvalue" = Var1/100) %>%
  mutate("Method" = Var2) %>%
  ggplot(aes_string(x = "pvalue", y = "value", color = "Method")) +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  xlab("P-value") +
  ylab("Power") +
  ggtitle(title)

}

plot_pvalue_hist <- function(pvalue_mat, title){
  melt(pvalue_mat) %>%
  mutate("Method" = Var2) %>%
  ggplot(aes(x = value, fill = Method)) +
  geom_histogram() +
  facet_wrap(~Method) +
  ggtitle(title)
}

plot_pvalue_scatter <- function(pvalue_mat, title){
  pvalue_mat %>%
  as.data.frame() %>%
  select(ANCOVA, strat = starts_with("Stratified")) %>%
  ggplot(aes(x = ANCOVA, y = strat)) + 
  geom_point() +
  xlim(0, 1) + ylim(0, 1) +
  ylab("Stratified Permutation") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  ggtitle(title)
}
```

@@

Before testing for different kinds of effects, we begin checking that the tests have the correct level.
We follow the procedure described above, using an effect size of $\gamma = 0$ at all strata and using standard normal errors.
To have the correct level means that the test rate of rejection at level $\alpha$ is $\alpha 100\%$.
In other words, the p-values are uniformly distributed and the power curve should coincide with the line with slope $1$ through the origin.
Figure \ref{fig:design_null_plots} demonstrates that this is the case.
If anything, the differenced stratified permutation test has fewer than $\alpha 100\%$ false positives when using level $\alpha$.

@@

```{r design_null_plots, fig.height = 4, fig.align="center"}
plot_power_curves(design0_power, "No Effect, Normal Errors")
plot_pvalue_hist(design0_pvalues, "No Effect, Normal Errors")
```

@@

```{r design_null_heavy_plots, fig.height = 4, fig.align="center"}
plot_power_curves(design00_power, "No Effect, Heavy-tailed Errors")
plot_pvalue_hist(design00_pvalues, "No Effect, Heavy-tailed Errors") 
```

@@

```{r design1_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design1_power, "Constant Additive Effect of 1, Normal Errors")
plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 1, Normal Errors")
```

@@

```{r design2_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design2_power, "Effect of 1 at a Single stratum, Normal Errors")
plot_pvalue_hist(design2_pvalues, "Effect of 1, Normal Errors")
```

@@

```{r design3_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design3_power, "Constant Additive Effect of 1, Heavy-tailed Errors")
plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 1, Heavy-tailed Errors")
```

@@

```{r design4_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design4_power, "Effect of 1 at a Single stratum, Heavy-tailed Errors") 
plot_pvalue_hist(design4_pvalues, "Effect of 1, Heavy-tailed Errors")
```

@@

```{r design5_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design5_power, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
plot_pvalue_hist(design5_pvalues, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
```

@@

```{r design6_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design6_power, "Constant Additive Effect of 1, Normal Errors")
plot_pvalue_hist(design6_pvalues, "Constant Additive Effect of 1, Normal Errors")
```

@@

```{r design7_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design7_power, "Effect of 1 at a Single stratum, Normal Errors")
plot_pvalue_hist(design7_pvalues, "Effect of 1, Normal Errors")
```

@@

```{r design8_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design8_power, "Constant Additive Effect of 1, Heteroskedastic Errors")
plot_pvalue_hist(design8_pvalues, "Constant Additive Effect of 1, Heteroskedastic Errors")
```

@@

```{r design9_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design9_power, "Effect of 1 at a Single stratum, Heteroskedastic Errors")
plot_pvalue_hist(design9_pvalues, "Effect of 1, Heteroskedastic Errors")
```

>>>>

plot_power_curves <- function(power_mat, title){
  melt(power_mat) %>% 
  mutate("pvalue" = Var1/100) %>%
  mutate("Method" = Var2) %>%
  ggplot(aes_string(x = "pvalue", y = "value", color = "Method")) +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  xlab("P-value") +
  ylab("Power") +
  ggtitle(title) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    title = element_text(size = 16),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    strip.text.x = element_text(size = 12)
  ) 
}
```

@@

Before testing for different kinds of effects, we begin checking that the tests have the correct level.
We follow the procedure described above, using an effect size of $\gamma = 0$ at all strata and using standard normal errors.
To have the correct level means that the test rate of rejection at level $\alpha$ is $100\alpha\%$.
In other words, the p-values are uniformly distributed and the power curve should coincide with the line with slope $1$ through the origin.
The figure below demonstrates that this is the case.
If anything, the differenced stratified permutation test has fewer than $100\alpha\%$ false positives when using level $\alpha$.

@@

```{r design_null_plots, fig.height = 4, fig.align="center"}
plot_power_curves(design0_power, "No Effect, Normal Errors")
```

@@

```{r design_null_heavy_plots, fig.height = 4, fig.align="center"}
plot_power_curves(design00_power, "No Effect, Heavy-tailed Errors")
```

@@

```{r design1_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design1_power, "Constant Additive Effect of 1 \n Normal Errors")
```

@@

```{r design2_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design2_power, "Effect of 1 at a Single stratum \n Normal Errors")
```

@@

```{r design3_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design3_power, "Constant Additive Effect of 1 \n Heavy-tailed Errors")
```

@@

```{r design4_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design4_power, "Effect of 1 at a Single stratum \n Heavy-tailed Errors") 
```

@@

```{r design5_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design5_power, "Constant Additive Effect of 1 \n Normal Errors \n Unpredictive baseline")
```

@@

```{r design6_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design6_power, "Constant Additive Effect of 1 \n Normal Errors")
```

@@

```{r design7_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design7_power, "Effect of 1 at a Single stratum \n Normal Errors")
```

@@

```{r design8_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design8_power, "Constant Additive Effect of 1 \n Heteroskedastic Errors")
```

@@

```{r design9_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design9_power, "Effect of 1 at a Single stratum \n Heteroskedastic Errors")
```
